Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-29 appendix.tex
minute changes to sec 3, 4, 5, appendix A; partial re-write of appendix B; new file for fig 1 not found, please update

\section{Derivation of the approximate equations}\label{adia_improve}
Here we detail the derivation of Eq. \ref{nearly_iso_explicit} used in the 
analytical derivations of \S\ref{analytical}.  Starting with Eqs.\ \ref{lin_mass} --- \ref{lin_energy},
we set $\Gamma = 1$ and $\zeta = 0$ for the vertically isothermal limit and the fully-radially-local 
approximation, respectively.  We  eliminate
the horizontal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{subequations}\label{A1}\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{d\ln{\rho}}{d z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{d\ln{\rho}}{dz}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}\end{subequations}
where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2,
\end{align} 
and we have replaced $\p/\p z$ by $d/dz$ for the fully radially local treatment.

Reduction to a single ODE requires vertical derivatives.  At this point we could use the 
low frequency and Keplerian approximations to set $D \rightarrow \kappa^2 \rightarrow \OmK^2$,
 and we would wind up with Eq. \ref{nearly_iso_explicit} more directly.
However to demonstrate that the order of approximation is irrelevant, we will retain
the vertical derivatives of the epicyclic frequency $\kappa$,
\begin{align}\label{dkappa2}
  \frac{d\kappa^2}{d z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{d\ln\rho}{d z}\frac{qc_s^2}{r^2} \left(
    \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{d\ln\rho}{d z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ increases monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$.

@@

We now eliminate $W,\, Q$ from Eqs.\ \ref{A1}, using  Eqs. \ref{chi} and \ref{dkappa2} to obtain  
\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z.
\end{align}

Now we make the low frequency and Keplerian
approximations, setting $D\to
\OmK^2$, to give  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3},
\end{align} 
in terms of dimensionless variables of Eq.\ \ref{nondim}.

Retaining terms to first order in the disk aspect ratio, $h \ll 1$, gives
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z.\label{vertiso_gov_nondim}
 \end{align}
Applying Eq.\ \ref{rhoisothin} for the density field gives the desired result, Eq. \ref{nearly_iso_explicit}.

@@

\section{Applicability of the  VGSB}\label{global_corr}
%In \S\ref{analytical} we effectively employed the vertically global
%shearing box \citep[VGSB,][]{mcnally14} to make analytic progress. 
The vertically global shearing box \citepalias[VGSB,][]{mcnally14} is an
extension of the standard shearing box \citep[SSB,][]{goldreich65} to
background shear flows that are height dependent. 
%As in the SSB, in
%the VGSB the equilibrium disk structure is assumed to be homogeneous
%in the radial ($x$) direction.  
Within the VGSB framework, the vertical shear $d\Omega/dz$
can be freely specified. We choose $d\Omega/dz$ to be that due to the
radial temperature gradient in the global disk
(Eq. \ref{vertical_shear}). However, explicit dependencies on the
radial disk structure are ignored elsewhere in the linearized equations
as described below. We show in the following sections that this
inconsistency only becomes important in the adiabatic limit, which is
not the relevant regime for the VSI.

%is only important when
%considering adiabatic limit, which is not the relevant regime for the
%VSI.

The background radial disk structure explicitly appears 
in the linearized mass, radial momentum and energy equations
(Eq. \ref{lin_mass}, \ref{lin_vx} and \ref{lin_energy}) as the terms
proportional to $\zeta$. Nominally $\zeta=1$.      
In the VGSB (and SSB) these terms are neglected by setting
$\zeta=0$. For a power-law disk, these radial gradients  
are $O(r^{-1})$, and they appear in comparison with terms of
$O(k_x)$. The neglected terms therefore have a relative magnitude of
$O(h/\khat)$, which is small for thin disks ($h\ll1$)
and/or small radial wavelengths ($\khat\gg 1$).

% The difference between the linear problem in the VGSB 
% and the radially-local approximation adopted in our numerical study
% is the absence of terms explicitly associated with the radial disk structure in
% the VGSB.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ and hence $\chi = 1/\gamma$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\zeta=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\upsilon}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI, which occurs 
for $\beta\ll1$.   %and small scale

@@

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\sigma\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider
$\beta\ll1$, then the VGSB framework is adequate for studying the
VSI.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    the linear problem, Eq. \ref{lin_mass}---\ref{lin_energy}, with $\zeta=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving 
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\zeta=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $(\gamma, \Gamma)=(1.4, 1.011)$ and
    $(p,q,h)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\upsilon}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii h q \hat{k} f(\zhat)
      }{1-\hat{\upsilon}^2} \right]W^\prime +
  \hat{\upsilon}^2\left(1+\frac{\hat{k}^2}{1-\hat{\upsilon}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = h^2q f(\hat{z})\OmK^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+h^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

>>>>

\section{Derivation of the approximate equations}\label{adia_improve}
Here we detail the derivation of Eq. \ref{nearly_iso_explicit} used in the 
analytical derivations of \S\ref{analytical}.  Starting with Eqs.\ \ref{lin_mass} --- \ref{lin_energy},
we set $\Gamma = 1$ and $\zeta = 0$ for the vertically isothermal limit and the fully-radially-local 
approximation, respectively.  We  eliminate
the horizontal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{subequations}\label{A1}\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{\p\ln{\rho}}{\p z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{\p\Omega^2}{\p z} -
    \frac{\p \ln{\rho}}{\p z}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{\p\ln{\rho}}{\p z}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}\end{subequations}
where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2.
\end{align} 
%

Reduction to a single ODE requires 
$\p_zD = \p_z\kappa^2$.  At this point we could apply the 
low frequency and Keplerian approximations to set $D \rightarrow
\kappa^2 \rightarrow \OmK^2$, then $D$ is vertically constant, 
and we can obtain Eq. \ref{nearly_iso_explicit} more directly.
However, to demonstrate that the order of approximation is irrelevant,
we will retain $\p_z\kappa^2$ initially. Using 
Eq. \ref{kap2_def}---\ref{vertical_shear},  
\begin{align}\label{dkappa2}
  \frac{\p \kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{d\ln\rho}{d z}\frac{qc_s^2}{r^2} \left(
    \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{d\ln\rho}{d z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ increases monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$.

@@

We  eliminate $W,\, Q$ from Eqs.\ \ref{A1}, using  Eqs. \ref{chi} and \ref{dkappa2} to obtain  
\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z, 
\end{align}
where we have replaced $\p/\p z$ by $d/dz$ for the
fully-radially-local treatment. Now we make the low frequency and Keplerian
approximations, setting $D\to 
\OmK^2$, to give  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3},
\end{align} 
in terms of dimensionless variables of Eq.\ \ref{nondim}.

Retaining terms to first order in the disk aspect ratio, $h \ll 1$, gives
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z.\label{vertiso_gov_nondim}
\end{align}
Approximating the density field by Eq.\ \ref{rhoisothin} then gives 
Eq. \ref{nearly_iso_explicit}.

@@

\section{Applicability of the fully-radially-local approximation}\label{global_corr}
In the fully-radially-local approximation, background radial
gradients are ignored except where it appears
implicitly in the expression for the vertical shear rate $\p_z\Omega$
(via Eq. \ref{vertical_shear}).  This is done by neglecting the terms proportional to $\zeta$
in Eq. \ref{lin_mass}, \ref{lin_vx} and \ref{lin_energy}, i.e. setting 
$\zeta=0$. (Nomially $\zeta=1$, which is used in our numerical study.)

For a power-law disk, the neglected radial gradients   
are $O(r^{-1})$, and they appear in comparison with terms of
$O(k_x)$. The neglected terms therefore have a relative magnitude of
$O(h/\khat)$, which is small for thin disks ($h\ll1$) 
and/or small radial wavelengths ($\khat\gg 1$).  We show in the
following sections that the fully-radially-local approximation only 
becomes invalid in the adiabatic limit, which is not the relevant
regime for the VSI.

We comment that this approximtaion is equivalent to 
that adopted in the vertically global shearing box formalism
\citepalias[VGSB,][]{mcnally14}, which is an 
extension of the standard shearing box \citep{goldreich65} to
background shear flows that are height dependent.

@@

\subsection{Spurious growth of adiabatic perturbations when
  $\zeta=0$}\label{analytic_adia}  
A limitation of the reduced model described in Appendix
\ref{adia_improve}, \S\ref{sec:simplified} and used in \S\ref{analytical}, is that it
cannot be employed for adiabatic flow when there is vertical shear, even
if $h/\khat\ll1$.  We explain this by setting  $\beta\to\infty$ and hence $\chi = 1/\gamma$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\zeta=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\upsilon}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because we have retained the global radial temperature 
gradient to obtain vertical shear, 
but have ignored it elsewhere in the linear problem (as well as the
background radial density gradient). Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI, which occurs 
for $\beta\ll1$.   %and small scale

@@

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the fully-radially-local
approximation, where $\zeta=0$, gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
expected behavior  ($\sigma\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider 
$\beta\ll1$, then setting $\zeta=0$ does not affect the VSI
significantly.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical ($\zeta=0$)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp}. 
    The `Numerical ($\zeta=0$)' curve is obtained from the numerical
    solution to Eq. \ref{lin_mass}---\ref{lin_energy} with
    $\zeta=0$,  which 
    neglects explicit dependencies on the radial disk structure.  
    The `Numerical
    ($\zeta=1$)' curve is obtained from Eq. \ref{lin_mass}---\ref{lin_energy} with
    $\zeta=1$, which accounts for the radial disk structure self-consistently.   
    The disk parameters are $(\gamma, \Gamma)=(1.4, 1.011)$ and
    $(p,q,h)=(-1.5,-1,0.05)$. The perturbation 
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the fully-radially local approximation
($\zeta=0$). In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\upsilon}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii h q \hat{k} f(\zhat)
      }{1-\hat{\upsilon}^2} \right]W^\prime +
  \hat{\upsilon}^2\left(1+\frac{\hat{k}^2}{1-\hat{\upsilon}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = h^2q f(\hat{z})\OmK^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+h^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

### 2015-04-29 application.tex
minute changes to sec 3, 4, 5, appendix A; partial re-write of appendix B; new file for fig 1 not found, please update

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
PPD do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one might consider an energy equation
with radiative cooling, 
\begin{align}\label{real_energy}
  % \rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
  \frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
    c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where $c_v$ is the heat capacity at constant
volume, 
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term to permit an equilibrium solution.

@@

\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear
problem by introducing higher order derivatives. We defer a proper
analysis of this problem to a follow-up study. Here, we are interested
in order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed 
by assuming vertical derivatives can be neglected in 
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
\citepalias[see also][]{barker15} where $\eta=k_\mathrm{rad}/\rho c_v$ is
the diffusion coefficient and 
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\OmK^{-1}}{\hat{\eta}\khat^2}, 
\end{align}

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our linear models, we estimate where in a
PPD do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one might consider an energy equation
with radiative cooling, 
\begin{align}\label{real_energy}
  % \rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
  \frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
    C_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where $C_v$ is the heat capacity at constant
volume, 
$\bm{F}$ is the radiative flux and $Q_+$ is a heating
term to permit an equilibrium solution.

@@

\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T C_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear
problem by introducing higher order derivatives. We defer a proper
analysis of this problem to a follow-up study. Here, we are interested
in order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed 
by assuming vertical derivatives can be neglected in 
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T C_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    C_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
\citepalias[see also][]{barker15} where $\eta=k_\mathrm{rad}/\rho C_v$ is
the diffusion coefficient and 
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 C_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\OmK^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
